---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Welcome to my Peer Review Assignment 1 for Reproducible Research class.

Make sure the working directory is set to the location of the activity.csv file.

First off the activity.csv file is read in.
```{r}
library("lattice")

datafile <- "activity.csv"
datafile <- paste0(getwd(), "/", datafile)

stepdata <- read.csv(datafile, header = TRUE, comment.char = "")
```

For future use the date and time fields are combined into a single POSIX format
and added as a column to the data.
Additionally a column is added to hold a logical, TRUE or FALSE, value if the
steps field is marked na.

```{r}
stepdata$datetime <- as.POSIXct(paste(stepdata$date, stepdata$interval%/%100,
                                      stepdata$interval%%100),
                                "%Y-%m-%d %H %M", tz = "GMT")
stepdata$isna <- is.na(stepdata$steps)
```

tapply is used to sum the total steps for each day, na's are removed before
summing.
allnas is a flag that a day consists entirely of na's.

The following histogram shows the frequency (how many times) a day had a step
total within the bar's range. Days that had all na's are not used in the
histogram since they their sum is 0, which is not a valid count of activity for
the day.

```{r}
dailysteps <- tapply(stepdata$steps, stepdata$date, sum, na.rm = TRUE)
allnas <- tapply(stepdata$isna, stepdata$date, all)

# don't use the days with all na's, since the sum will be 0 rather then na
hist(dailysteps[!allnas], breaks = 10,
     main = "Histogram of Total Steps per Day", xlab = "Total Daily Steps")

cat("The mean steps per day is: ", mean(dailysteps))
cat("The median steps per day is: ", median(dailysteps))

rm(allnas)               # housekeeping
```

tapply is once again used to calculate the mean number of steps taken for each
time interval across all the days. The maximum activity is noted both in the
graph and on a line of output.

```{r}
steps <- tapply(stepdata$steps, stepdata$interval, mean, na.rm = TRUE)

time <- as.integer(names(steps))
timePOSIX <- as.POSIXct(paste(time %/% 100, time %% 100, sep = " "),
                   "%H %M", tz = "GMT")
timesteps <- data.frame(steps, time, timePOSIX)

rm(steps, time, timePOSIX)         # housekeeping

plot(timesteps$timePOSIX, timesteps$steps, type = "l", xaxt = "n",
     xlab = "Time of Day (hh:mm)", ylab = "5 Minute Step Average",
     main = "Average Daily Activity")
axis(1, at = as.POSIXct(c("00","6","12","18","24"), "%H", tz = "GMT"),
     labels = c("00:00","6:00","12:00","18:00","24:00"))

maxsteps <- which.max(timesteps$steps)
timeofmax <- timesteps$timePOSIX[maxsteps]
maxsteps <- timesteps$steps[maxsteps]

points(timeofmax, maxsteps, pch=1, col = "mediumblue")
text(timeofmax, maxsteps,
     paste0("(",format(timeofmax, "%H:%M"), ", ", round(maxsteps, 0), ")" ),
     pos=4 , cex = 0.67, col = "mediumblue")

cat("On average at ", format(timeofmax, "%H:%M"), " the maximum number of steps, ", round(maxsteps, 0), ", occurred.", sep = "")

```

A seperate data.frame is created from the original. The mean daily activity
calculated above for each time interval is used to replace missing values in
the original data.
The code assumes that all 288 (24 * 60 / 5) time intervals are present for each
day. This assumption appears to hold for this activity file since
17568 / 288 = 61 even.

As above the number of steps for each day is again computed including the 
imputed data and a histogram generated.

```{r}
 
cat("There are", sum(stepdata$isna), "missing step values in the dataset")

# to fill in missing data use the mean for that time step.
# timesteps holds one days info, so will be repeated for each day in stepdata
imputeddata <- stepdata
imputeddata$imputedsteps <- timesteps$steps
imputeddata$imputedsteps <- ifelse(imputeddata$isna, 
                                   imputeddata$imputedsteps, imputeddata$steps)
# 
dailyimputsteps <- tapply(imputeddata$imputedsteps, imputeddata$date,
                          sum, na.rm = FALSE)
hist(dailyimputsteps, breaks = 10,
     main = "Histogram of Total Steps per Day", xlab = "Total Daily Steps")

cat("The mean steps per day is: ", mean(dailyimputsteps))
cat("The median steps per day is: ", median(dailyimputsteps))

```

Days are then factored as a weekend or weekday.
A lattice plot comparing the weekend and weekday activity is then presented.

```{r}
imputeddata$weekday <- weekdays(imputeddata$datetime)
imputeddata$weekday <- ifelse(imputeddata$weekday %in% c("Saturday", "Sunday"),
                              "weekend", "weekday")
imputeddata$weekday <- as.factor(imputeddata$weekday)

weekdaysteps <- tapply(imputeddata$imputedsteps[imputeddata$weekday == "weekday"],
                       imputeddata$interval[imputeddata$weekday == "weekday"],
                       mean)
weekendsteps <- tapply(imputeddata$imputedsteps[imputeddata$weekday == "weekend"],
                       imputeddata$interval[imputeddata$weekday == "weekend"],
                       mean)
timeimputed <- data.frame(timesteps$timePOSIX, weekdaysteps)
timeimputed$dayofweek <- as.factor("weekday")
colnames(timeimputed)[2] <- "steps"
timeimputed2 <- data.frame(timesteps$timePOSIX, weekendsteps)
timeimputed2$dayofweek <- as.factor("weekend")
colnames(timeimputed2)[2] <- "steps"
timeimputed <- rbind(timeimputed, timeimputed2)

colnames(timeimputed)[1] <- "timePOSIX"

xyplot(steps ~ timePOSIX | dayofweek, timeimputed, type = "l", 
       scales = list(at = as.POSIXct(c("00","6","12","18","24"), "%H", tz = "GMT"),
                 labels = c("00:00","6:00","12:00","18:00","24:00")),
       layout = c(1,2),
       xlab = "Time of Day (hh:mm)", ylab = "Average Steps")
```